---
title: "Titanic Survival Analysis"
author: "xipecast"
date: "2026"
output: html_document
---

## Overview
Analysis of 891 Titanic passengers to determine survival factors.

- **Rows**: 891 passengers
- **Columns**: 12 variables
- **Target**: Survival (0=Died, 1=Survived)

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_minimal(base_size = 14))

titanic <- read_csv("/home/boss/Documents/myDocuments/ProjectsOnLenovo/xipecast-analyst-portfolio/titanic-analysis/data/titanic.csv")

titanic_clean <- titanic %>%
  mutate(Age = ifelse(is.na(Age), median(Age, na.rm = TRUE), Age)) %>%
  mutate(Embarked = ifelse(is.na(Embarked), "S", Embarked)) %>%
  select(-Cabin)
```

## 1. Overall Survival
```{r survival-count}
titanic_clean %>%
  mutate(Survived = ifelse(Survived == 1, "Survived", "Died")) %>%
  ggplot(aes(x = Survived, fill = Survived)) +
  geom_bar() +
  labs(title = "Titanic Survival Count", x = "Status", y = "Count")
```

## 2. Survival by Gender
```{r survival-gender}
titanic_clean %>%
  mutate(Survived = ifelse(Survived == 1, "Survived", "Died")) %>%
  ggplot(aes(x = Sex, fill = Survived)) +
  geom_bar(position = "dodge") +
  labs(title = "Survival by Gender", x = "Gender", y = "Count")
```

### Gender Survival Rates
```{r gender-rates}
titanic_clean %>%
  group_by(Sex) %>%
  summarise(
    Total = n(),
    Survived = sum(Survived),
    Rate = paste0(round(Survived/Total * 100, 1), "%")
  )
```

## 3. Survival by Class
```{r survival-class}
titanic_clean %>%
  mutate(Survived = ifelse(Survived == 1, "Survived", "Died")) %>%
  ggplot(aes(x = factor(Pclass), fill = Survived)) +
  geom_bar(position = "dodge") +
  labs(title = "Survival by Class", x = "Class (1=First, 2=Second, 3=Third)", y = "Count")
```

### Class Survival Rates
```{r class-rates}
titanic_clean %>%
  group_by(Pclass) %>%
  summarise(
    Total = n(),
    Survived = sum(Survived),
    Rate = paste0(round(Survived/Total * 100, 1), "%")
  )
```

## 4. Combined Gender + Class
```{r combined}
titanic_clean %>%
  group_by(Sex, Pclass) %>%
  summarise(
    Total = n(),
    Survived = sum(Survived),
    Rate = paste0(round(Survived/Total * 100, 1), "%")
  )
```

```{r combined-plot}
titanic_clean %>%
  group_by(Sex, Pclass) %>%
  summarise(
    survival_rate = round(sum(Survived)/n() * 100, 1)
  ) %>%
  ggplot(aes(x = factor(Pclass), y = survival_rate, fill = Sex)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = paste0(survival_rate, "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5) +
  labs(title = "Survival Rate by Gender & Class",
       subtitle = "Gender mattered more than class",
       x = "Class", y = "Survival Rate (%)")
```

## 5. Survival by Age
```{r survival-age}
titanic_clean %>%
  mutate(Survived = ifelse(Survived == 1, "Survived", "Died")) %>%
  ggplot(aes(x = Age, fill = Survived)) +
  geom_histogram(bins = 30, position = "dodge") +
  labs(title = "Survival by Age", x = "Age", y = "Count")
```

## 6. Fare vs Survival
```{r fare-survival}
titanic_clean %>%
  mutate(Survived = ifelse(Survived == 1, "Survived", "Died")) %>%
  ggplot(aes(x = Survived, y = Fare, fill = Survived)) +
  geom_boxplot() +
  labs(title = "Fare vs Survival", x = "Status", y = "Fare Paid")
```

## Key Findings
| Factor | Finding |
|--------|---------|
| **Gender** | Women survived at 74.2% vs men at 18.9% |
| **Class** | 1st class 63% vs 3rd class 24.2% |
| **Combined** | Rich women = 96.8% survival, Poor men = 13.5% |
| **Age** | Mean age deaths (30.6) vs survived (28.3) |
| **Fare** | Higher fare = better survival odds |

## Conclusion
> A passenger's best chance of survival was being **female and first class.**
> Poor males in 3rd class had only a **13.5% chance** of survival.
> Gender was a stronger predictor than class or age.